\hl{Identifying Effective Motivational Interviewing Communication Sequences Using Automated Pattern Analysis}

### 2018-06-28 jhir/latex-manuscript/manuscript.tex
prepared final copy for the minor revision

HMMs are widely used for the analysis of sequence data due to their ability to model \hl{long-range dependencies} between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution. Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state. HMMs were originally proposed for speech recognition~\cite{rabiner1989tutorial}, in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}. For example, \hl{an HMM-based} classifier was \hl{applied to Doppler} ultrasound imaging data to extract features \hl{from} the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}. In another study, HMM was used to capture important
characteristics of protein families~\cite{amrhein2003client}. In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states \hl{may correspond to a patients' underlying} motivational state during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.

>>>>

HMMs are widely used for the analysis of sequence data due to their ability to model \hl{long-range dependencies} between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution. Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state. HMMs were originally proposed for speech recognition~\cite{rabiner1989tutorial}, in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}. For example, \hl{an HMM-based} classifier was \hl{applied to Doppler} ultrasound imaging data to extract features \hl{from} the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}. In another study, HMM was used to capture important
characteristics of protein families~\cite{amrhein2003client}. In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states \hl{may correspond to patients' underlying} motivational state during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.
